### 2013-12-03 paper/main.tex
cleaned up diagram and some text, minor

The engine model used here is based work
by Butts and Sivashankar\autocite{532315} which was derived from
the work by Powell and Cook\autocite{4789342}.
The configuration is a modern 4.6L V-8 gas engine.
A single control input representing the idle air bypass valve

To simplify analysis the linearized model is used as shown in
Figure \ref{fig:lem}.

@@

\begin{tikzpicture}[>=triangle 60, scale=0.7,
	node distance=14mm, auto]

@@

	\draw [->] (-20mm,0) node[above]{$\Delta \text{control}$} -- (gain1.west);
	\draw [->] (gain1) -- (sum1);
	\draw [->] (sum1) -- (gain2);
	\draw [->] (gain2) -- (sum2);
	\draw [->] (sum2) -- (gain3);
	\draw [->] (gain3.east) -- ++(20mm,0) node[above]{$\Delta \text{rpm}$};
	\draw [->] ($(gain3.east) + (10mm,0) $) |- (gain4.east);
	\draw [->] (gain4.west) -| (sum1.south);
	\draw [<-] (sum2.north) -- ++(0,10mm) node[above]{$\Delta \text{torque}$};
\end{tikzpicture}

@@

This model takes two inputs: a torque, and a idle control signal.
When the torque is greater than zero it will oppose the rotation of
the engine causing it to slow down.
The idle control signal is some fraction of unity.
This fraction corresponds to a pulse width modulated idle control valve
which is at a minimum near zero and at a maximum near unity.
Often the duty cycle range is in a range from 0\% to 100\% which
corresponds to 0 to 1 (unity).

Because all the inputs and outputs are defined as deltas ($\Delta$)
this model cannot be used directly with typical control systems
which use steady state values.
It is possible convert these deltas to steady state equivalents.
Figure \ref{fig:sdtf} shows the transfer function to convert steady
state values to delta values.
Figure \ref{fig:dstf} shows the transfer function to delta values
to steady state values.

% {{{ fig:sdtf
\begin{figure}[!htbp]

\begin{center}
\begin{tikzpicture}[>=triangle 60]
\node[gain] (gain1) [] {$1 - z^{-1}$};
\draw [->] (-20mm,0) node[above,yshift=1mm]{$X$} -- (gain1.west);
\draw [->] (gain1.east) -- (20mm,0) node[above,yshift=1mm]{$\Delta X$};
\end{tikzpicture}
\end{center}

\caption{The $Z$ transform used to accumulate the input and convert
a steady state input to delta output.
Its derivation is given in Appendix \ref{app:cdelta}.}
\label{fig:sdtf}

@@

\caption{The $Z$ transform used to convert delta input to a steady
state output.
Its derivation is given in Appendix \ref{app:dsderiv}.}
\label{fig:dstf}

@@

% TODO - reduced steady state system

In order to build a controller the model needs to be simplified
but the presence of two inputs complicates matters.
To resolve this issue the torque can be set zero.
Then it can be simplified by recognizing that it matches
the well known form shown in Figure \ref{fig:dd} which
has the transfer function in Equation \ref{eq:dd}.

@@

First taking the parts from the engine model (Figure \ref{fig:lem}).

@@

\begin{samepage}
Then substituting them in to Equation \ref{eq:dd} and simplifying
results in Equation \ref{eq:lems}.
\begin{align}
	\frac{y}{u} &= \frac{(1.699) (8.5683) (2.98) (1)}
		{(z^2 - 0.9025z)(z - 0.9354)(1) + (8.5683)(2.98)(0.00093)} \notag \\
	\Aboxed{ \frac{y}{u} &= \frac{43.38}{z^3 - 1.838 z^2 + 0.8442 z + 0.02375}}
		\label{eq:lems}
\end{align}
\end{samepage}

\begin{samepage}
It is still necessary to address the delta inputs and outputs.
This can be resolved by placing the conversion transforms,
from Figure \ref{fig:sdtf} and \ref{fig:dstf}, on either
end of the engine model transfer function as shown in Figure \ref{fig:nodelta}.
A beneficial side effect becomes apparent in this form.
The transform that converts from steady state to a delta cancels
with the transform that converts from a delta to steady state
\footnote{In the more general case, when torque is non zero,
the transforms would not cancel.}.
Therefore the final transform is still Equation \ref{eq:lems}.
An example of the output response with no control is shown in
Figure \ref{fig:em_plot}.
\end{samepage}

@@

% {{{ fig:em_plot
\begin{figure}[htbp!]
\begin{center}
\includegraphics[scale=0.7]{../octave/engine_model_plot}
\end{center}
\caption{Output response of engine model with no control.
The control can range from zero to one.
Here the control is constant at 0.65 which resulted in a stable
rpm near 900.\protect\footnotemark}
\label{fig:em_plot}
\end{figure}
\footnotetext{The Matlab source for this and the other engine model plots
are in Appendix \ref{app:em_matlab}.}
% }}}

It is not apparent from Figure \ref{fig:em_plot} exactly how long
this process takes.
The $x$ axis represents ignition events, not time.
The rpm can be converted to time per ignition event by examining
the units as shown in Equation \ref{eq:rpmtime}.

@@

Applying this equation results in the response shown
in Figure \ref{fig:rpmtime_plot}.
It can be seen that the rise time over $0.5$ second and
that the overshoot is close to $40\%$.
The stabilization time is also several seconds.
Clearly its performance could be improved with the application
of a controller.

@@

The response from zero rpm is not a realistic measurement of performance
because zero rpm is impossibly slow.
A typical idle rpm range is from 600 to 1000 rpm.

Figure \ref{fig:em_shift_plot2} shows the response for a step input
with control inputs chosen to reach 600 and 1000 rpm
\footnote{The control input to stable rpm values are given in
Appendix \ref{app:contrpm}}.
This is a more realistic representation of typical operation.
The rise time is near $0.2$ seconds and the overshoot is $15\%$.
And the stabilization time is almost one second.
Its behavior can still be improved with a controller.

>>>>

The engine model used here is based work
by Butts and Sivashankar\autocite{532315} which was derived from
the work by Powell and Cook\autocite{4789342}.
The configuration is a modern 4.6L V-8 gas engine.
To simplify this analysis the linearized model was used as shown in
Figure \ref{fig:lem}.

@@

\begin{tikzpicture}[>=triangle 60, scale=0.6,
	node distance=14mm, auto]

@@

	\draw [->] (-30mm,0) node[above,xshift=2mm,yshift=1mm]{$\Delta \text{control}$}
				-- (gain1.west);
	\draw [->] (gain1) -- (sum1);
	\draw [->] (sum1) -- (gain2);
	\draw [->] (gain2) -- (sum2);
	\draw [->] (sum2) -- (gain3);
	\draw [->] (gain3.east) -- ++(30mm,0) node[above,xshift=-5mm,yshift=1mm]{$\Delta \text{rpm}$};
	\draw [->] ($(gain3.east) + (10mm,0) $) |- (gain4.east);
	\draw [->] (gain4.west) -| (sum1.south);
	\draw [<-] (sum2.north) -- ++(0,10mm) node[above]{$\Delta \text{torque}$};
\end{tikzpicture}

@@

This model takes two inputs: a torque, and a idle control signal.
When the torque is greater than zero it will oppose the rotation of
the engine causing it to slow down.
The idle control signal is some fraction of unity.
This fraction corresponds to a pulse width modulated idle control valve
which is at a minimum near zero and at a maximum near unity.

Because all the inputs and outputs are defined as deltas ($\Delta$)
this model cannot be used directly with typical control systems
which expect steady state values.
It is possible to convert these deltas to steady state equivalents.
Figure \ref{fig:sdtf} shows the transfer function to convert steady
state values to delta values
\footnote{The derivation of the steady state to delta conversion
is given in Appendix \ref{app:cdelta}}.
Figure \ref{fig:dstf} shows the transfer function to delta values
to steady state values
\footnote{The derivation of the delta to steady state conversion
is given in Appendix \ref{app:dsderiv}.}.

% {{{ fig:sdtf
\begin{figure}[!htbp]
\begin{center}
\begin{tikzpicture}[>=triangle 60]
\node[gain] (gain1) [] {$1 - z^{-1}$};
\draw [->] (-20mm,0) node[above,yshift=1mm]{$X$} -- (gain1.west);
\draw [->] (gain1.east) -- (20mm,0) node[above,yshift=1mm]{$\Delta X$};
\end{tikzpicture}
\end{center}
\caption[xyz]{The $Z$ transform used to accumulate the input and convert
a steady state input to delta output.}
\label{fig:sdtf}
\end{figure}

@@

\caption{The $Z$ transform used to convert delta input to a steady
state output.}
\label{fig:dstf}

@@

In order to construct a controller the model (Figure \ref{fig:lem})
needs to be simplified in to a transfer function.
However the prescence of two inputs, control and torque, complicates matters.
To resolve this issue the torque can be set zero.
Then it can be simplified by recognizing that it matches
the well known form shown in Figure \ref{fig:dd} which
has the transfer function in Equation \ref{eq:dd}.

@@

To simply first gather the parts from the engine model (Figure \ref{fig:lem}).

@@

\begin{samepage}
Then substitute them in to Equation \ref{eq:dd} and simplify.
The result is Equation \ref{eq:lems}.
\begin{align}
	\frac{y}{u} &= \frac{(1.699) (8.5683) (2.98) (1)}
		{(z^2 - 0.9025z)(z - 0.9354)(1) + (8.5683)(2.98)(0.00093)} \notag \\
	\Aboxed{ \frac{y}{u} &= \frac{43.38}{z^3 - 1.838 z^2 + 0.8442 z + 0.02375}}
		\label{eq:lems}
\end{align}
\end{samepage}

\begin{samepage}
It is still necessary to address the delta inputs and outputs.
This can be resolved by placing the conversion transforms
(from Figure \ref{fig:sdtf} and \ref{fig:dstf}) on either
end of the engine model as shown in Figure \ref{fig:nodelta}.
A beneficial side effect becomes apparent in this form.
The transform that converts from steady state to a delta cancels
with the transform that converts from a delta to steady state
\footnote{In the more general case, when torque is not zero,
the transforms wouldn't cancel.}.
Therefore the final transform is still Equation \ref{eq:lems}.
An example of the output response with no control is shown in
Figure \ref{fig:em_plot}.
The Matlab source code for this plot and others are
in Appendix \ref{app:em_matlab}.
\end{samepage}

@@

% {{{ fig:em_plot
\begin{figure}[htbp!]
\begin{center}
\includegraphics[scale=0.7]{../octave/engine_model_plot}
\end{center}
\caption{Output response of engine model with no control.
The control can range from zero to one.
Here the control is constant at 0.65 which resulted in a stable
rpm near 900.}
\label{fig:em_plot}
\end{figure}
% }}}

It is not apparent from Figure \ref{fig:em_plot} exactly how long
this process takes because the $x$ axis represents ignition events
instead of time.
However the rpm can be converted to time per ignition event by examining
the units as shown in Equation \ref{eq:rpmtime}.

@@

Applying Equation \ref{eq:rpmtime} results in the response shown
in Figure \ref{fig:rpmtime_plot}.
It can be seen that the rise time is over $0.5$ second and
that the overshoot is close to $40\%$.
The stabilization time is also several seconds.
Clearly its performance could be improved with the application
of a controller.

@@

The response from zero rpm is not a realistic measurement of performance
because zero rpm is impossibly slow.
A typical idle rpm range is from 600 to 1000 rpm.
Figure \ref{fig:em_shift_plot2} shows the response for a step input
with control inputs chosen to reach 600 and 1000 rpm
\footnote{The control input to stable rpm values are given in
Appendix \ref{app:contrpm}}.
This is a more realistic representation of typical operation.
The rise time is near $0.2$ seconds and the overshoot is $15\%$.
And the stabilization time is almost one second.
Its behavior can still be improved with a controller.
